Rotation Periods in Low-Mass Binary Stars: The Impact of Tidal Torques and Magnetic Braking

### 2018-12-07 Paper/sync.tex
Updates from ShareLaTeX

Stellar rotation period (P$_{rot}$) studies have produced numerous insights into the angular momentum evolution of low-mass stars, but previous investigations have focused on single stars even though roughly half of Sun-like stars are in stellar binaries. Here, we examine how tidal torques, stellar evolution, and magnetic braking shape the P$_{rot}$ evolution of low-mass stellar binaries by performing Monte Carlo simulations of coupled stellar-tidal evolution. We find that tides can modify stellar rotations in binaries with orbital periods (P$_{orb}$) up to 100 days.  At short P$_{orb}$, many stellar binaries tidally-lock into synchronous rotation, naturally producing a population of fast rotators (P$_{rot} < 10$ d) across all stellar masses as is observed in the \textit{Kepler} field, a feature that cannot be reproduced by single-star-only models of magnetic braking.  Many binaries with longer P$_{orb}$ can tidally-lock, fixing the stellar P$_{rot}$ to the binary P$_{orb}$, causing P$_{rot}$ to not be a valid proxy for age in all cases, i.e. gyrochronology methods must be applied carefully. We compare our model predictions with \kepler eclipsing binary (EB) rotation periods and find that our predictions are in excellent agreement with observations, with the ``Constant Time Lag" equilibrium tidal best matching observations.  We show that even for short P$_{orb}$ binaries, tidal torques can balance magnetic braking, producing a population of subsynchronous rotators similar what \citet{Lurie2017} observed in their sample of \kepler EBs. By comparing observations of \kepler EBs with our simulations, we constrain low-mass stellar tidal time lags to $\tau = 0.19^{+0.43}_{-0.16}$ s.

@@

Angular momentum evolution in low-mass (M$\lsim 1$ M$_{\odot}$) short-period (P$_{orb} \lsim 10$ d) stellar binaries is dominated by tides.  Tidal torques drive secular changes in the binary orbit and stellar spins, eventually circularizing the orbit and synchronizing the stellar spins in the long-term \citep{Counselman1973}. Orbital circularization is ubiquitous in short-period binaries, owing to the tidal torque's strong semi-major axis dependence, with both theoretical \citep[e.g.][]{Zahn1989} and observational \citep[e.g.][]{Meibom2005,Mazeh2008,Lurie2017} studies finding that most binaries with P$_{orb} \lsim 10$ d are circularized. For short-period binaries, tidal torques work quickly on ${\sim}100$ Myr timescales, as \citet{Zahn1989} finds that the orbit of solar twin binaries circularizes during the stellar pre-main sequence.  Observations by \citet{Meibom2005} support this picture as they find short-period binaries in the ${\sim}150$ Myr old cluster M35 tend to have circular orbits.

Tides impart a significant signature in the long-term angular momentum evolution for binary stars, especially so for the stellar spins. Tidal torques drive stellar spins towards the tidally-locked state in which the stellar P$_{rot}$ is equal to the equilibrium rotation period (P$_{eq}$) predicted by tidal models, with a familiar example of this effect being spin-orbit synchronization where P$_{rot} = $ P$_{eq} = $ P$_{orb}$.  Tidal-locking occurs much earlier than orbital circularization as the tidal-locking timescale is estimated to be $2-3$ orders of magnitude less than the circularization timescale \citep{Zahn1989,Witte2002,Mazeh2008}, and as a result, tidal-locking is expected for binaries with P$_{orb} \lsim$ 20 d \citep[e.g.][]{Meibom2006,Mazeh2008,Zahn2008,Meibom2015}.

@@

We simulate coupled stellar-tidal evolution for stellar binaries using an improved version of the coupled stellar-tidal evolution model presented in \citet{Fleming2018}.  We implement our model in the open-source code VPLanet\footnote{VPLanet is publicly available
at \href{https://github.com/VirtualPlanetaryLaboratory/vplanet}{{https://github.com/VirtualPlanetaryLaboratory/vplanet}}.} \citep[][Barnes et al., in prep]{Barnes2016,vplanet2018}.  We integrate all model equations (see $\S$~\ref{sec:methods:stellar} and $\S$~\ref{sec:methods:eqtide}) using using the $4^{th}$ order Runge-Kutta scheme with adaptive timestepping described in \citet{Fleming2018}.

@@

We examine stellar angular momentum evolution in low-mass binaries by performing a Monte Carlo study of two sets of 10,000 stellar binaries, one modeled using the CPL model and the other using the CTL formalism.  We simulate both stars' spin evolution but mainly consider the P$_{rot}$ evolution for the primary, i.e. more massive star in binaries, as it is observationally easier to measure a P$_{rot}$ on the more massive, and hence brighter, star \citep[e.g.][]{Meibom2006,Lurie2017}. For each simulation, we sample the primary's mass uniformly over $[0.1, 1]$ M$_{\odot}$. Following \citet{Matt2015}, we uniformly sample the log$_{10}$ of P$_{rot}$ over [$0.8,15$] days, a distribution that approximates the P$_{rot}$ distribution of young stars in the ${\sim}2$ Myr old ONC.  We compute the secondary star's mass by uniformly sampling the mass ratio over [$0.1, 1$] following observations of mass ratios in low-mass binaries \citep{Raghavan2010,Moe2018}. Given the inherent uncertainty in and complexity of the formation of binaries \citep[e.g.][]{Bonnell1994,Bate2000,Bate2002,Moe2018} and the potential for dynamical processing via tides or stellar close encounters \citep[e.g.][]{Mardling2001,Hurley2002,Ivanova2005,Meibom2005}, we take an agnostic approach to the initial orbital configuration by uniformly randomly sampling the initial eccentricity ($e$) over [$0.0,0.3$] and uniformly sampling the initial P$_{orb}$ over [$3,100$] d. Although the CTL model is applicable for $e > 0.3$, the CPL model is not, so we restrict $e \leq 0.3$ to allow us to compare both models. We do not consider P$_{orb} < 3$ d as these binaries are likely to have a tertiary companion \citep{Tokovinin2006} which can significantly impact the inner binaries' dynamical evolution \citep[e.g.][]{Munoz2015,Martin2015b,Hamers2016,Moe2018}.

Values for stellar tidal $Q$s and $\tau$s for low-mass stars are highly uncertain, likely depend on complex viscous evolution within the stars \citep{Ogilvie2007}, can differ for stars of the same spectral class \citep{Barker2009}, and can vary as a function of stellar age \citep{Bolmont2016}, likely due to low-mass star's evolving convective regions where the tidal dissipation predominantly occurs \citep{Zahn2008}. Typical values of $Q$ and $\tau$ for Sun-like stars are estimated to be of order $Q \approx 10^6$ and $\tau \approx 0.1$ s, respectfully \citep[e.g.][]{Meibom2005,Ogilvie2007,Jackson2009}, however a wide range of values exist in the literature.  Therefore, we consider a wide range of tidal parameters by sampling stellar tidal Qs loguniformly over $[10^4,10^7]$ and $\tau$ loguniformly over $[10^{-2},1]$ s.  All stars have an initial age of 5 Myr unless stated otherwise as by this time, the gaseous protoplanterary circumbinary disk that can impart significant dynamical changes in the system \citep[e.g.][]{Fleming2017} would likely have dissipated \citep{Haisch2001}.

@@

Spin-orbit synchronization should not be assumed for short P$_{orb}$ binaries. In Fig.~\ref{fig:eqPerShortPorb}, we plot the P$_{rot}$ evolution for a P$_{orb} = 5$ d binary for various tidal dissipation parameters and find that subsynchronous rotation occurs in the general case of weak tidal torques, $Q>10^7$ or $\tau < 0.1$ s in these cases, and is not restricted to long P$_{orb}$ binaries.  Moreover, short P$_{orb}$ subsynchronous rotators can eventually tidally-lock via the mechanism described above where tidal torques gradually strengthen relative to magnetic braking. Short P$_{orb}$ subsynchronous binaries exist in nature, such as some \kepler EBs (\citet{Lurie2017}, see $\S$~\ref{sec:kepler} for further discussion), Kepler-47 \citep{Orosz2012} and a Ruprecht 147 eclipsing binary \citep{Torres2018}, suggesting that this theoretical observation is real.  We explore this effect further in $\S$~\ref{sec:subsync} and describe how we can use it to constrain the tidal parameters $Q$ and $\tau$.

@@

In Fig.~\ref{fig:lurie7}, we display P$_{orb}/$P$_{rot}$ as a function of P$_{orb}$ for both the CPL and CTL models where each simulation was integrated to an age uniformly sampled over $1-7$ Gyr, consistent with ages of \kepler field stars \citep{Chaplin2014}.

- Consult Lurie's paper for CPL vs CTL (e.g. why not 3:2 population-> use CTL!) and for discussing sunsync population. stars prob not in 3:2 resonance since they likely dont have a fixed solid body shape as I assume, but perhaps in a decoupled core envelope model (bouvier?) envelope could get locked into resonance

- compare results to Lurie section 4.4.2 discussion of dependence with eccentricity \citet{Burkart2014}

- lurie finds that low mass ratio binaries are more likely to be subsync for Porb less than 10 days, makes sense since lower mass secondary means weaker tidal torques

@@

@@

\subsubsection{P$_{orb} < 10$ d: The Subsynchronous Population} \label{sec:subsync}

@@

Although Lurie+2017 attributes differential rotation to the production of the
subsynchronous population, we find that coupled stellar-tidal evolution naturally
produces the population. Compare with Lurie+2017 sample for 2 $<$ Porb $<$ 10 days:
$72\%$ in [0.92,1.2] and $15\%$ in [0.84, 0.92].

@@

@@

@@

@@

- consider dynamical tidal effects (Zahn 1975)?

>>>>

 Tidal torques dominate the angular momentum evolution of short-period, low-mass stellar binaries, but few previous studies have examined tides in binaries out to intermediate orbital periods (P$_{orb}$). Here, we examine how tidal torques, stellar evolution, and magnetic braking shape the rotation period (P$_{rot}$) evolution of low-mass stellar binaries with P$_{orb} \leq 100$ d by performing Monte Carlo simulations of coupled stellar-tidal evolution. We find that tides can modify stellar rotations in binaries with orbital periods (P$_{orb}$) up to 100 days.  At short P$_{orb}$, many stellar binaries tidally-lock into synchronous rotation, naturally producing a population of fast rotators (P$_{rot} < 10$ d) as is observed in the \textit{Kepler} field, a feature that cannot be reproduced by single-star-only models of magnetic braking.  Many binaries with longer P$_{orb}$ can tidally-lock, fixing the stellar P$_{rot}$ to the binary P$_{orb}$, causing P$_{rot}$ to not be a valid proxy for age in all cases, i.e. gyrochronology methods must be applied carefully. We compare our model predictions with \kepler eclipsing binary (EB) rotation periods and find that our predictions are in excellent agreement with observations, with the ``Constant Time Lag" equilibrium tidal model best matching observations. We recommend observational tests to discriminate between equilibrium tidal models for stars.  We show that when tidal torques are weak, even in short P$_{orb}$ binaries, tidal torques can overpower magnetic braking, producing a long-lasting population of subsynchronous rotators similar what \citet{Lurie2017} observed in their sample of \kepler EBs. By comparing observations of \kepler EBs with our simulations, we loosely constrain low-mass stellar tidal time lags to $\tau = 0.19^{+0.43}_{-0.16}$ s.

@@

Angular momentum evolution in low-mass (M$\lsim 1$ M$_{\odot}$) short-period (P$_{orb} \lsim 10$ d) stellar binaries is dominated by tides.  Tidal torques drive secular changes in the binary orbit and stellar spins, eventually circularizing the orbit and synchronizing the stellar spins in the long-term \citep{Counselman1973}. Orbital circularization is ubiquitous in short-period binaries, owing to the tidal torque's strong semi-major axis dependence, with both theoretical \citep[e.g.][]{Zahn1989,Claret1995} and observational \citep[e.g.][]{Meibom2005,Mazeh2008,Lurie2017} studies finding that most binaries with P$_{orb} \lsim 10$ d are circularized. For short-period binaries, tidal torques work quickly on ${\sim}100$ Myr timescales, as \citet{Zahn1989} finds that the orbit of solar twin binaries circularizes during the stellar pre-main sequence.  Observations by \citet{Meibom2005} support this picture as they find short-period binaries in the ${\sim}150$ Myr old cluster M35 tend to have circular orbits.

Tides impart a significant signature in the long-term angular momentum evolution for binary stars, especially so for the stellar spins. Tidal torques drive stellar spins towards the tidally-locked state in which the stellar P$_{rot}$ is equal to the equilibrium rotation period (P$_{eq}$) predicted by tidal models, with a familiar example of this effect being spin-orbit synchronization where P$_{rot} = $ P$_{eq} = $ P$_{orb}$.  Tidal-locking occurs much earlier than orbital circularization as the tidal-locking timescale is estimated to be $2-3$ orders of magnitude less than the circularization timescale \citep{Zahn1989,Witte2002,Mazeh2008}, and as a result, tidal-locking is expected for binaries with P$_{orb} \lsim$ 20 d \citep[e.g.][]{Levato1974,Meibom2006,Mazeh2008,Zahn2008,Meibom2015}.

@@

We simulate coupled stellar-tidal evolution for low-mass binaries using an improved version of the coupled stellar-tidal evolution model presented in \citet{Fleming2018}.  We implement our model in the open-source code VPLanet\footnote{VPLanet is publicly available
at \href{https://github.com/VirtualPlanetaryLaboratory/vplanet}{{https://github.com/VirtualPlanetaryLaboratory/vplanet}}.} \citep[][Barnes et al., in prep]{Barnes2016,vplanet2018}.  We integrate all model equations (see $\S$~\ref{sec:methods:stellar} and $\S$~\ref{sec:methods:eqtide}) using using the $4^{th}$ order Runge-Kutta scheme with adaptive timestepping described in \citet{Fleming2018}.

@@

We examine stellar angular momentum evolution in low-mass binaries by performing a Monte Carlo study of two sets of 10,000 stellar binaries, one modeled using the CPL model and the other using the CTL formalism.  We simulate both stars' spin evolution but mainly consider the P$_{rot}$ evolution for the primary, i.e. more massive star in binaries, as it is observationally easier to measure a P$_{rot}$ on the more massive, and hence brighter, star \citep[e.g.][]{Meibom2006,Lurie2017}. For each simulation, we sample the primary's mass uniformly over $[0.1, 1]$ M$_{\odot}$. Following \citet{Matt2015}, we uniformly sample the log$_{10}$ of P$_{rot}$ over [$0.8,15$] days, a distribution that approximates the P$_{rot}$ distribution of young stars in the ${\sim}2$ Myr old ONC.  We compute the secondary star's mass by uniformly sampling the mass ratio over [$0.1, 1$] following observations of mass ratios in low-mass binaries \citep{Raghavan2010,Moe2018}. Given the inherent uncertainty in and complexity of the formation of binaries \citep[e.g.][]{Bonnell1994,Bate2000,Bate2002,Moe2018} and the potential for dynamical processing via tides or stellar close encounters \citep[e.g.][]{Mardling2001,Hurley2002,Ivanova2005,Meibom2005}, we take an agnostic approach to the initial orbital configuration by uniformly randomly sampling the initial eccentricity ($e$) over [$0.0,0.3$] and uniformly sampling the initial P$_{orb}$ over [$3,100$] d. Although the CTL model is applicable for $e > 0.3$, the CPL model is not, so we restrict $e \leq 0.3$ to allow us to compare both models. We do not consider P$_{orb} < 3$ d as these binaries are likely to have a tertiary companion \citep{Tokovinin2006} which can significantly impact the inner binary's dynamical evolution \citep[e.g.][]{Munoz2015,Martin2015b,Hamers2016,Moe2018}.

Values for stellar tidal $Q$s and $\tau$s for low-mass stars are highly uncertain, likely depend on complex viscous evolution within the stars \citep{Ogilvie2007}, can differ for stars of the same spectral class \citep{Barker2009}, and can vary as a function of stellar mass or age \citep{Bolmont2016,vanEylen2016}, likely due to low-mass star's evolving convective regions where the tidal dissipation predominantly occurs \citep{Zahn2008}. Typical values of $Q$ and $\tau$ for Sun-like stars are estimated to be of order $Q \approx 10^6$ and $\tau \approx 0.1$ s, respectfully \citep[e.g.][]{Meibom2005,Ogilvie2007,Jackson2008}, however a wide range of values exist in the literature.  Therefore, we consider a wide range of tidal parameters by sampling stellar tidal $Q$s loguniformly over $[10^4,10^7]$ and $\tau$ loguniformly over $[10^{-2},1]$ s.  Note that in general, there is no expression to compute $Q$ as a function of $\tau$, or vice versa, except in some special cases where approximations exist, e.g. Eqn. (2) from \citet{Heller2011}. All stars have an initial age of 5 Myr unless stated otherwise as by this time, the gaseous protoplanterary circumbinary disk, that can drive significant dynamical evolution in the binary \citep[e.g.][]{Fleming2017}, would likely have dissipated \citep{Haisch2001}.

@@

Spin-orbit synchronization should not be assumed for short P$_{orb}$ binaries. In Fig.~\ref{fig:eqPerShortPorb}, we plot the P$_{rot}$ evolution for a P$_{orb} = 5$ d binary for various tidal dissipation parameters and find that subsynchronous rotation occurs in the general case of weak tidal torques, $Q>10^7$ or $\tau < 0.1$ s in these cases, and is not restricted to long P$_{orb}$ binaries.  Moreover, short P$_{orb}$ subsynchronous rotators can eventually tidally-lock via the mechanism described above where tidal torques gradually strengthen relative to magnetic braking. Short P$_{orb}$ subsynchronous binaries exist in nature, such as some \kepler EBs (\citet{Lurie2017}, see $\S$~\ref{sec:kepler} for further discussion), Kepler-47 \citep{Orosz2012} and EPIC 219394517 \citep{Torres2018}, suggesting that this theoretical observation is real.  We explore this effect further in $\S$~\ref{sec:subsync} and describe how we can use it to constrain the tidal parameters $Q$ and $\tau$.

@@

In Fig.~\ref{fig:lurie7}, we display P$_{orb}/$P$_{rot}$ as a function of P$_{orb}$ for both the CPL and CTL models where each simulation was integrated to an age uniformly sampled over $1-7$ Gyr, consistent with ages of \kepler field stars \citep{Chaplin2014}.  Both models reproduce most trends in the \citet{Lurie2017} data: the large cluster of synchronized rotators with P$_{orb} \lsim 20$ d, a population of subsynchronous rotators with P$_{orb} \lsim 20$ d, and the sparse cluster of supersynchronous rotators at all P$_{orb}$ that correspond to young stars that have yet to significantly spin down due to magnetic braking or tides.  The lower limit seen in both panels is a line of nearly constant P$_{rot} \approx 60$ d set by how much a star can spin down via magnetic braking in the weak tides regime over 7 Gyr, the longest age considered in our simulations.

@@

The predictions of the CPL model fail to reproduce several key aspects of that data.  First, the CPL model cannot reproduce the cluster of supersynchronous rotators with P$_{orb}/$P$_{rot} \lsim 1.1$ for P$_{orb} < 10$ d and instead predicts that all sypersynchronous rotators lie on the line P$_{orb}/$P$_{rot} \lsim 1.5$, a prediction that is entirely inconsistent with the data as no obvious spin-orbit commensurablities, aside from 1:1 synchronization, is present in the data likely owing to the fact that stellar convective envelopes lack a fixed shape, making resonant coupling difficult unless it occurs with internal gravity or pressure modes \citep{Burkart2014,Lurie2017}.  The CPL model can produce the subsynchronous rotators observed for $10 \lsim$ P$_{orb} \lsim 20$ d, but fails to produce the prominent cluster of subsynchronous rotators at P$_{orb}/$P$_{rot} \approx 0.9$ for P$_{orb} < 10$ d observed by \citet{Lurie2017}.  The CTL model, however, excels where the CPL models fails and naturally produces both populations, suggesting that the CTL model better describes tidal interactions in low-mass binaries.  We examine this point further by comparing results from our CTL model with the \citet{Lurie2017} data where the data is most complete, P$_{orb} < 10$ d, in $\S$~\ref{sec:subsync}.

@@

\citet{Claret1995} find that supersynchronous rotation their sample of eccentric binaries are in good agreement with the predictions of tidal models.

presence of 3:2 cluster would be strong evidence in favor of CPL model.  Can use presence/absence of this feature as a strong discriminators between tidal models.  additionally, CTL predicts more subsync clusters that could be used for model descriptions. below, say how CPL model has locked binaries at long Porb, ctl doesn't above 20 d so that's another model descriimination.  say each time that CTL passes, CPL doesn't according to lurie data.  In general, talk about implications and use flowery language

\subsubsection{P$_{orb} < 10$ d According to the CTL Model} \label{sec:subsync}

@@

- lurie finds that low mass ratio binaries are more likely to be subsync for Porb less than 10 days, makes sense since lower mass secondary means weaker tidal torques.

Although Lurie+2017 attributes differential rotation to the production of the
subsynchronous population, we find that coupled stellar-tidal evolution naturally
produces the population. it's probably a combination of both effects. Compare with Lurie+2017 sample for 2 $<$ Porb $<$ 10 days:
$72\%$ in [0.92,1.2] and $15\%$ in [0.84, 0.92].

@@

XXX: say how our quoted $Q$ and $\tau$ ranges produced binaries consistent with what \kepler saw, so not only is our model a reasonable choice to model stellar-tidal evolution, but also our quoted tidal parameter ranges seem to bracket what stars actually have.  We refine it in the below analysis

@@

XXX:

-hammer implications for gyrochronology, fast rotators in kepler field, compare to \citet{Simonian2018}

- \citet{Abt2004} find evidence for tides modifying P$_{rot}$ in binaries with P$_{orb}$ up to 500 d.

@@

- talk about how we need more binary observations with Prot, Porb, and ecc, especially at longer orbital periods, to more reliably discriminate between tidal models.  At low eccentricity, it does matter (sync vs supersync in \kepler) as opposed to what \citet{Leconte2010} recommends

@@
